Whose Truth is it Anyway? An Experiment on Annotation Bias

### 2023-06-12 report/draft.Rmd
integration NLP text

Generating large data-sets has become one of the main drivers of progress in natural language understanding. In studies of political communication, the most familiar annotation tasks involve identifying basic concepts. This often involve noting the topic of the text, the position of the actor or the tone of the text. A recent study [@van2021validity] demonstrates that using crowd-coding platforms is a good way to collect such large data-sets. However, having only a few workers annotate the majority of text of interest has raised concerns about data diversity and models' ability to generalize beyond the crowd-workers: In a series of experiments, @geva2019we show that often models do not generalize well to annotations from annotators that did not contribute to the training set, suggesting that annotator bias should be monitored during data-set creation. One such potential bias, especially in times of increasing polarization [@iyengar2019origins; @gidron2019toward; @boxell2022cross], is based on ideological position of the annotator. Given that annotators on crows-coding platforms tend to be younger and hold more liberal values than the general public [@clifford2015samples; @huff2015these], this could potentially hamper the data diversity and generalizability of the model. An additional reason to monitor the annotators ideological position as a potential source of annotator bias is that a recent study in NLP showed that experiential factors influence the consistency of how political ideologies are perceived [@shen2021sounds]. Their finding challenges the "ground-truth" assumption we as researcher make that a position for example is either left-leaning or right-wing leaning. People with different ideological backgrounds might experience that position differently. This challenges us. We are interested in the effect of e.g. elite communication. To study this, we allow for heterogeneous treatment effects in experimental work. This indicates that we often do not assume that the treatment, often using text, has the same effect for different partisans. Yet, at the same time, we forget or ignore that knowledge when creating large data-sets for our machine learning models.

The field of Natural Language Processing often works on automatically classifying texts on labels of concepts such as stance, sentiment, and political orientation. These models are trained with training data, which is annotated by human annotators. Often, this process has a final step where disagreements and differences in annotations are levelled by aggregating, averaging, or in other ways coming to a consensus on one label for one example, which is then seen as "ground truth"[@aroyo2015]. Differences from this ground truth label are seen as errors that need to be removed or sorted out. Models then learn to predict labels for new examples based on this ground truth. However, since several years there is some discussion on how realistic it is to have one label, especially for subjective or complex concepts and/or texts with multiple interpretations. @aroyo2015 describes succinctly how the predominant annotation procedures for classification models run into myths such as "disagreement is bad" and "one annotation is enough". @plank2014learning adds to this that underlying ambiguity and linguistic complexity should be considered for disagreement in annotations: not all linguistic examples are created equal. Disagreement even occurs in seemingly objective tasks such as Part of Speech tagging [@fornaciari2021, @plank2014learning].

Another aspect to consider is that disagreement can be informative for the concept under measure - sometimes agreement can be used to validate hypotheses about how universal the perceptions of such concepts are [@sommerauer]. Additional doubts on smoothing out disagreement in annotation have focussed on the lack of diversity when only annotating with one label or annotator, leading to a homogeneity especially in subjective and social tasks [@geva2019we] such as hatespeech detection or political affiliation classification. The question then is: whose perspective is being recorded in these datasets, and then later in the models trained on these datasets? Or, as [@geva2019we] asked: are we modelling the task, or the annotator?

Several tasks are intrinsically societal, with answers that differ based on the make-up of the community of annotators. The answer of the annotators in turn influence how machine learning examples classify new models. For instance, hate speech detection and abuse detection is one NLP task where the race and gender of annotators influences annotators and model performance[@gordon2022, @Larimore, @waseem2016]. Language is inherently connected to society and culture: @shenDarling analyze sentiment analysis, and find that human annotators lead to certain perspectives on sentiment being recorded, and that notably that African American English dialects are often misunderstood by such models.

Most recently, new annotation paradigms have gone one step further: maybe we are annotating the annotator rather than the task, and maybe that is actually beneficial to the models. @pavlick2019 find that for the logical coherence task Natural Language Inference, annotators have several valid interpretations that are not reflected in one ground truth label. They call for new training paradigms that can reflect "the full range of possible human inferences"[@pavlick2019, p. 688]. Recent approaches in NLP have sought to explicitly incorporate disagreement and diversity in training data annotations. [@röttger2021] introduces the idea of an explicitly subjective annotation paradigm existing in addition to one focussed on one label and "ground truth". Such a subjective annotation paradigm can be used for purposes where the goal is finding diverse perspectives on the task or concepts, and for models to model more accurately how humans interpret a task or text. Additionally, "perspectivism" [@basile2021toward] is a paradigm and research agenda where different perspectives are explicitly incorporated in the training data, and used by models to provide more human-like classifications. Another paradigm is "jury learning" [@gordon2022], in which machine learning models do not learn to replicate one specific ground truth, but are trained with different annotator juries to reflect the judgement of different populations. In both approaches, demographic and other individual aspects of the annotator are explicitly mentioned and highlighted as having an influence on classification performance, but is used as an asset rather than as an error.

@@

As noted by @plank2014, some linguistic examples are ambiguous, and open to multiple interpretations even for seemingly objective tasks such as part of speech tagging, where annotators have to distinguish parts of speech such as nouns and verbs. A complex concept such as political ideology is much more likely to lead to multiple interpretations. We call such sentences with more possible interpretations and less explicit standpoints "underspecified". A lack of explicitness in the annotated text is one of the main causes of the disagreements in earlier literature. @thornjakobsen2022 deduce that annotator bias coms from a process nown as the affect heuristic [@slovic2007 ]: making a secision based on the emotional response related to your own personal attitude towards the discussed topic, especially when the text is relatively ambiguous.

@@

If these biases exists, how can we alleviate them?

There have been several previous approaches to solve biased annotations, especially where it concerns political or societal aspects. @geva2019we introduce an approach where training set annotators are separated from annotators annotating datasets that evaluate the models, to ensure the evaluation is not simply accurate at replicating the original annotators, but can generalize to new annotators' judgements. However, these approaches are not aimed at reducing biases during the training set annotation procedure. Other approaches are aimed at leveraging multiple perspectives - but this is not useful when one wants one label to learn from.

Very woke hot take that I am unsure whether to include and also where: "\"The problem is that framing arbitrary representations in data as \"bias\" misses the political character of datasets: there is no neutral data and no apolitical standpoint from where we can call out bias [23]. Datasets are always \"a worldview\" [26] and, as such, data always remains biased.\" (Miceli et al., 2022, p. 5)"

For the Austrian National Elections, @ennser2018impact already demonstrated that showing party labels impact annotators' assessment of the party position. So, one solution is masking:

@@

Think about perspectivism for another solution!

>>>>

Generating large data-sets has become one of the main drivers of progress in natural language understanding. In studies of political communication, the most familiar annotation tasks involve identifying basic concepts. This often involve noting the topic of the text, the position of the actor or the tone of the text. A recent study [@van2021validity] demonstrates that using crowd-coding platforms is a good way to collect such large data-sets. However, having only a few workers annotate the majority of text of interest has raised concerns about data diversity and models' ability to generalize beyond the crowd-workers: In a series of experiments, @geva2019we show that often models do not generalize well to annotations from annotators that did not contribute to the training set, suggesting that annotator bias should be monitored during data-set creation. One such potential bias, especially in times of increasing polarization [@iyengar2019origins; @gidron2019toward; @boxell2022cross], is based on ideological position of the annotator. Given that annotators on crows-coding platforms tend to be younger and hold more liberal values than the general public [@clifford2015samples; @huff2015these], this could potentially hamper the data diversity and generalizability of the model. An additional reason to monitor the annotators ideological position as a potential source of annotator bias is that a recent study in NLP showed that experiential factors influence the consistency of how political ideologies are perceived [@shen2021sounds]. Their finding challenges the "ground-truth" assumption we as researcher make that a position for example is either left-leaning or right-wing leaning. People with different ideological backgrounds might experience that position differently. 
This challenges our way of data collection.
We are interested in the effect of e.g. elite communication. To study this, we allow for heterogeneous treatment effects in experimental work. This indicates that we often do not assume that the treatment, often using text, has the same effect for different partisans. Yet, at the same time, we forget or ignore that knowledge when creating large data-sets for our machine learning models.

The field of Natural Language Processing often works on automatically classifying texts on labels of concepts such as stance, sentiment, and political orientation. These models are trained on data created by human annotators. Often, this process has a final step where disagreements and differences in annotations are leveled by aggregating, averaging, or in other ways coming to a consensus on one label for one example, which is then seen as "ground truth" [@aroyo2015]. 
Differences from this ground truth label are seen as errors that need to be removed or sorted out. 
Models then learn to predict labels for new examples based on this ground truth. However, since several years there is some discussion on how realistic it is to have one label, especially for subjective or complex concepts and/or texts with multiple interpretations. @aroyo2015 describes succinctly how the predominant annotation procedures for classification models run into myths such as "disagreement is bad" and "one annotation is enough". @plank2014learning adds to this that underlying ambiguity and linguistic complexity should be considered for disagreement in annotations: not all linguistic examples are created equal. Disagreement even occurs in seemingly objective tasks such as Part of Speech tagging [@fornaciari2021, @plank2014learning].

Another aspect to consider is that disagreement can be informative for the concept under measure - sometimes agreement can be used to validate hypotheses about how universal the perceptions of such concepts are [@sommerauer2020penguin]. 
Additional doubts on smoothing out disagreement in annotation have focussed on the lack of diversity when only annotating with one label or annotator, leading to a homogeneity especially in subjective and social tasks [@geva2019we] such as hatespeech detection or political affiliation classification.
The question then is: _Whose perspective is being recorded in these datasets, and then later in the models trained on these datasets?_ 
Framing arbitrary representations in data as \"bias\" misses the political character of datasets: There is no neutral data and no apolitical standpoint from where we can call out bias. Datasets are always \"a worldview\" [26] and, as such, data always remains biased.\" (Miceli et al., 2022, p. 5).
This is key to social scientist in general, and those studying political text in particular, since several tasks of interest are intrinsically societal, with answers that differ based on the make-up of the worldview of annotators. 
The answer of the annotators in turn influence how machine learning examples classify new models. 
For instance, hate speech detection and abuse detection is one NLP task where the race and gender of annotators influences annotators and model performance[@gordon2022, @Larimore, @waseem2016]. 
Language is inherently connected to society and culture: @shenDarling analyze sentiment analysis, and find that human annotators lead to certain perspectives on sentiment being recorded, and that notably that African American English dialects are often misunderstood by such models.

Most recently, new annotation paradigms have gone one step further by asking whether we  modelling the task, or the annotator [@geva2019we]?
@pavlick2019 find that for the logical coherence task Natural Language Inference, annotators have several valid interpretations that are not reflected in one ground truth label. They call for new training paradigms that can reflect "the full range of possible human inferences"[@pavlick2019, p. 688]. Recent approaches in NLP have sought to explicitly incorporate disagreement and diversity in training data annotations. [@röttger2021] introduces the idea of an explicitly subjective annotation paradigm existing in addition to one focussed on one label and "ground truth". Such a subjective annotation paradigm can be used for purposes where the goal is finding diverse perspectives on the task or concepts, and for models to model more accurately how humans interpret a task or text. Additionally, "perspectivism" [@basile2021toward] is a paradigm and research agenda where different perspectives are explicitly incorporated in the training data, and used by models to provide more human-like classifications. Another paradigm is "jury learning" [@gordon2022], in which machine learning models do not learn to replicate one specific ground truth, but are trained with different annotator juries to reflect the judgement of different populations. In both approaches, demographic and other individual aspects of the annotator are explicitly mentioned and highlighted as having an influence on classification performance, but is used as an asset rather than as an error.

@@

As noted by @plank2014, some linguistic examples are ambiguous, and open to multiple interpretations even for seemingly objective tasks such as part of speech tagging, where annotators have to distinguish parts of speech such as nouns and verbs. A complex concept such as political ideology is much more likely to lead to multiple interpretations. We call such sentences with more possible interpretations and less explicit standpoints "underspecified". A lack of explicitness in the annotated text is one of the main causes of the disagreements in earlier literature. @thornjakobsen2022 deduce that annotator bias coms from a process nown as the affect heuristic [@slovic2007 ]: making a secision based on the emotional response related to your own personal attitude towards the discussed topic, especially when the text is relatively ambiguous.
To test whether people with different ideological backgrounds as well as their political knowledge might experience that position differently, challenging our ground-truth assumption in data annotation, we propose the following hypotheses:

@@

_If these biases exists, how can we alleviate them?_
There have been several previous approaches to solve biased annotations, especially where it concerns political or societal aspects. @geva2019we introduce an approach where training set annotators are separated from annotators annotating datasets that evaluate the models, to ensure the evaluation is not simply accurate at replicating the original annotators, but can generalize to new annotators' judgements. However, these approaches are not aimed at reducing biases during the training set annotation procedure. Other approaches are aimed at leveraging multiple perspectives - but this is not useful when one wants one label to learn from.
For the Austrian National Elections, @ennser2018impact already demonstrated that showing party labels impact annotators' assessment of the party position. So, one solution is masking:

@@
